# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

Here we go again...
```{r}
###library dplyr###

library(dplyr)
library(GGally)
library(ggplot2)
#read dataset from local directory

l14 <- read.csv("C:/Users/Jukke/OneDrive/Työpöytä/IODS/IODS-project/data/ln14_analysis.csv") %>% select(gender, Age, attitude, deep, stra, surf, Points)

```

graphical overview

```{r}
p <- ggpairs(l14, mapping = aes(col = gender, alpha = 0.3,), lower = list(combo = wrap("facethist", bins = 20)))
# draw the plot
p

```

explanation


```{r}
l14_sum <- summary(l14)
l14_sum
```

Gender
```{r}
l14_gen <- count(l14, gender)
l14_gen
```

Attitude and Points

```{r}
plot_ap <- ggplot(l14, aes(x = attitude, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Attitude vs. Points") + labs(x = "Attitude")
plot_ap
```

Strategy and Points
```{r}
plot_sp <- ggplot(l14, aes(x = stra, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Strategy vs. Points") + labs(x = "Strategy")
plot_sp
```

Surface and Points
```{r}
plot_fp <- ggplot(l14, aes(x = surf, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Surface Learning vs. Points") + labs(x = "Surface")
plot_fp
```

Multiple Regression

```{r}
multiple_regression <- lm(Points ~ attitude + stra + surf, data = l14)
summary(multiple_regression)
```

Multiple Regression Redone
```{r}
regression <- lm(Points ~ attitude, data = l14)
summary(regression)
```

Diagnostic Plots
```{r}
dp <- plot(regression, which = c(1,2,5))

dp
```
