# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

```{r}
date()
```

Here we go again...
```{r}
###library dplyr###

library(dplyr)
library(GGally)
library(ggplot2)
library(kableExtra)
#read dataset from local directory

l14 <- read.csv("C:/Users/Jukke/OneDrive/Työpöytä/IODS/IODS-project/data/ln14_analysis.csv") %>% select(gender, Age, attitude, deep, stra, surf, Points)

```

# 2

## 2.1 Graphical Overview
A graphical overview in Figure 1 displays the datasets 7 variables (gender, age, attitude, deep learning, strategic learning, surface learning and exam points) in relation to each other.

```{r}
p <- ggpairs(l14, mapping = aes(alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p

```

## 2.2 Variables

This section discusses the variables by exploring the dataset with the `summary` function and a `count` of gender. 
```{r}
l14_sum <- summary(l14)
l14_sum
l14_gen <- count(l14, gender)

kable(l14_gen, format = "html", caption = "Students' Gender")
```

**Age**: Respondents are aged between 17 and 55 with the median age being 22.  
**Gender**: Binary Male/Female with a distribution of 56/110  
**Learning approaches**: Deep, surface and strategic learning approaches scored by means discussed in previous section.  
**Attitude**: Students' attitude score  
**Points**: Exam points with a range of 7-33 and a mean of 22.72

# 3

## 3.1 Selecting Explanatory Variables

This section will investigate the relation between the students' exam points and their deep and surface learning approaches as well as their attitude. The objective is to find out if the exploratory variables (Attitude, Surface learning, Deep learning) are associated in a systematic way to students' exam points. 

### 3.1.1 Attitude and Exam Points

```{r}
plot_ap <- ggplot(l14, aes(x = attitude, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Attitude vs. Points") + labs(x = "Attitude")
plot_ap
```

### 3.1.2 Deep Learning and Exam Points
```{r}
plot_dp <- ggplot(l14, aes(x = deep, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Deep Learning vs. Points") + labs(x = "Deep Learning")
plot_dp
```

### 3.1.3 Surface Learning and Exam Points
```{r}
plot_sp <- ggplot(l14, aes(x = surf, y = Points, col = gender)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Student's Surface Learning vs. Points") + labs(x = "Surface Learning")
plot_sp
```

## 3.2 Multiple Regression

### 3.2.1 Overview of Variables
```{r}
l14_b <- select(l14, gender, attitude, surf, deep, Points)
p2 <- ggpairs(l14_b, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p2
```
  
### 3.2.2 Multiple Regression Model


```{r}
multiple_regression <- lm(Points ~ attitude + deep + surf, data = l14)
mr <- summary(multiple_regression)
print(mr)
```
We fail to find statistical significance in the p-values for deep learning (p < 0.23) and surface learning (p < 0.19). However, attitude seems statistically significant. In the next section let us examine the linear regression of exam points and attitude.

## Model Changed to Simple Linear Regression
```{r}
regression <- lm(Points ~ attitude, data = l14)
summary(regression)
```


Diagnostic Plots
```{r}
dp <- plot(regression, which = c(1,2,5))

dp
```
